\large{Vipassana for Hackers} \\ \Huge{The Proposal} \\ \large\textit{Version 0.1}

### 2024-06-26 the-proposal/the-proposal.tex
review/update abstract

\begin{abstract}
Vipassana meditation claims to ultimately explore the entire field of
mind and matter, with the goal of total liberation from suffering. Implicit
within these claims is a complete understanding of human consciousness. These are
difficult claims to prove or disprove because the time commitment required to research the
technique orders on multiple decades --- if not multiple generations.
This is complicated by the
fact that the time commitment is demanded of both the researcher \textit{and the
subject}. Until now, research on meditation of all kinds has covered only one of
two fields:
(1) controlled experiments which must inherently rely on superficial data
gathered from beginners \cite{https://doi.org/10.1002/jclp.20761,
  https://doi.org/10.1037/a0018441,
  https://doi.org/10.1093/scan/nss054,
  https://doi.org/10.31231/osf.io/hzv65}
and (2) observational exploratory research of monks --- expert
meditators who have dedicated their lives to the practice.
\cite{https://doi.org/10.1073/pnas.0407401101,
https://doi.org/10.1371/journal.pone.0073417}

I propose bridging this
gap by submitting myself to the middle ground. While remaining a layperson, I will
commit to a high ratio of waking medition hours. Individually
and internally, I will conduct qualitative research into the consequences of
Vipassana meditation and the nature of consciousness while collectively and
externally pledging myself as a subject for long-term quantitative
studies with a broader community of researchers.

@@

Because Vipassana \footnote{For the remainder of this research proposal,
  ``Vipassana'' will always refer to the specific technique ``Vipassana as taught by S.N. Goenka in the
  tradition of Sayagyi U Ba Khin'' and lineage-identical instruction, such as that
  taught by Ledi Sayadaw (in writing) or other students of U Ba Khin contemporary
  to S.N. Goenka, unless noted otherwise.} is globally standardized, it affords
researchers with definitive
solutions to the difficulties presented by varying teaching methodologies. In
exchange for this, the difficulty of long-term study is compounded by the intrinsic
seriousness demanded of a Vipassana practitioner: the Pāli concept of \textit{ātāpī
  sampajāno satimā} (continuous piercing awareness of constantly
changing bodily sensation \cite{goenka1999discourses}) is not only a
requirement of serious Vipassana practice but could very well act as a surrogate
description of the practice itself. This seriousness poses obvious difficulties and
it is the intention of my study to begin breaking ground in solutions to those
difficulties.

>>>>

\begin{abstract}
Vipassana meditation claims to ultimately explore the entire field of
mind and matter, with the goal of total liberation from suffering. Implicit
within this claim is a complete understanding of human consciousness.
This is a difficult claim to prove or disprove because the time commitment
required to research the technique orders on multiple decades
--- if not multiple generations.
Only a longitudinal study has any meaning.
This is complicated by the fact that the time commitment is demanded of
both the researcher \textit{and the subject}.
Until now, research on meditation of all kinds has covered only one of two fields:
(1) controlled experiments which must inherently rely on superficial data
gathered from beginners \cite{https://doi.org/10.1002/jclp.20761,
  https://doi.org/10.1037/a0018441,
  https://doi.org/10.1093/scan/nss054,
  https://doi.org/10.31231/osf.io/hzv65}
and (2) observational exploratory research of monks --- expert
meditators who have dedicated their lives to the practice.
\cite{https://doi.org/10.1073/pnas.0407401101,
https://doi.org/10.1371/journal.pone.0073417}

I propose bridging this
gap by submitting myself to the middle ground. While remaining a layperson, I will
commit to a high ratio of waking medition hours.
Individually and internally, I will conduct qualitative research into the consequences of
Vipassana meditation and the nature of consciousness.
Collectively and externally I will pledge myself as a subject for long-term quantitative
studies with a broader community of researchers.

@@

Because Vipassana \footnote{For the remainder of this research proposal,
  ``Vipassana'' will always refer to the specific technique ``Vipassana as taught by S.N. Goenka in the
  tradition of Sayagyi U Ba Khin'' and lineage-identical instruction, such as that
  taught by Ledi Sayadaw (in writing) or other students of U Ba Khin contemporary
  to S.N. Goenka, unless noted otherwise.} is globally standardized, it affords
researchers with definitive
solutions to the difficulties presented by varying teaching methodologies. In
exchange for this, the difficulty of long-term study is compounded by the intrinsic
seriousness demanded of a Vipassana practitioner: the Pāli concept of \textit{ātāpī
  sampajāno satimā} (continuous piercing awareness of constantly
changing bodily sensation \cite{goenka1999discourses, analayo2003satipatthana})
is not only a
requirement of serious Vipassana practice but could very well act as a surrogate
description of the practice itself. This seriousness poses obvious difficulties and
it is the intention of my study to begin breaking ground in solutions to those
difficulties.
